---
title: "NEON download"
author: "Kari Norman"
date: "10/15/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(neonUtilities)
library(piggyback)
```

Download data
```{r}
beetles_raw <- loadByProduct("DP1.10022.001", check.size = FALSE)
usethis::use_data(beetles_raw)
```

Grab necessary columns from different tables. Important column info:
* sampleID: unique ID for each combination of domainID, siteID, plotID, trapID, and collectDate (unique ID for bet_fielddata table)
* subsampleID: unique ID for each group of individuals, multiple per sample (unique ID for bet_sorting); unclear why there are muliple entries for the same classification
* individualID: unique ID for each individual identified by taxonomists (unique ID for bet_parataxonimist table)

There are multiple levels of taxonomic identification, need to decide which level of rigor we care about (NEON id often just to family, taxonomist ID to species but only for a subset, DNA ID for even smaller subset).
```{r}
data <- beetles_raw$bet_fielddata %>%
  filter(sampleCollected == "Y") %>% #there's an entry for every trap, whether or not they got samples, only want ones with samples
  select(sampleID, domainID, siteID, plotID, trapID, collectDate) %>%
  left_join(beetles_raw$bet_sorting %>% 
              filter(sampleType %in% c("carabid", "other carabid")) %>% #only want carabid samples, not bycatch
              select(sampleID, subsampleID, taxonID, scientificName, taxonRank, individualCount), 
            by = "sampleID") %>%
  filter(!is.na(subsampleID)) %>% #even though they were marked a sampled, some collection times don't acutally have any samples 
  left_join(beetles_raw$bet_parataxonomistID %>% select(subsampleID, individualID, taxonID, scientificName, taxonRank), by = "subsampleID") %>%
    mutate(year = lubridate::year(collectDate),
                month = lubridate::month(collectDate),
                day = lubridate::day(collectDate))
```

How many sites?
How many plots?
How many samples per?
```{r}
data %>%
  select(siteID, plotID, trapID, collectDate, year, month) %>%
  group_by(siteID, plotID, year, month) %>% 
  count() %>% 
  View()
```

What is the distribution of number of samples per plot across years?

What is the distribution of number of samples per trap
```{r}
sample_freq <- data %>%
  select(siteID, plotID, trapID, year, month, collectDate) %>%
  group_by(siteID, plotID, trapID, year) %>% 
  summarise(num_samples = n_distinct(collectDate))

hist(sample_freq$num_samples)
```



