---
title: "codyn"
author: "Kari Norman"
date: "1/27/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(codyn)
library(tidyverse) 
library(lubridate)
library(ggpubr)
```

See Avolio et al 2019 for latest package info

Comparing RACs:
* species richness (lenght of the RAC) 
* evenness (slope of the RAC, flatter = greater evenness)
* rank changes/reordering
* species loss
* species gain

Beetles
```{r}
load("data/codyn_beetles_inter.rda")
load("data/codyn_beetles_intra.rda")

#this data has already excluded sites that were not sampled according to our cutoff (four years of data with at least 3 replicates in a year)

beet_change_inter <- RAC_change(codyn_beetles_inter, 
           time.var = "year", species.var = "scientificName", abundance.var = "cpue", replicate.var = "siteID")

beet_change_intra <- RAC_change(codyn_beetles_intra %>% rename(bout = collectDate),
           time.var = "bout", species.var = "scientificName", abundance.var = "cpue", replicate.var = "siteID") %>%
  #remove bout comparisons that cross years
  filter(year(bout) == year(bout2))
```

Mammals
```{r}
#sites that meet sampling frequency criteria 
mam_inc_sites <- readRDS(file="data/Mammals/NEONsites_tempreplic_4yrs.rds") %>% 
  rename(siteID = SiteID)

mam_inter <- readRDS("~/Documents/Projects/temporalNEON/data/Mammals/DATA_mammals_abs-abund-inter.rds") %>%
  pivot_longer(-c(siteID, year), names_to = "scientificName", values_to = "abundance") %>%
  filter(siteID %in% mam_inc_sites$siteID)

mam_intra <- readRDS("~/Documents/Projects/temporalNEON/data/Mammals/DATA_mammals_abs-abund-intra.rds") %>%
  pivot_longer(-c(siteID, year, month, bout), names_to = "scientificName", values_to = "abundance") %>%
  filter(siteID %in% mam_inc_sites$siteID)


mam_change_inter <- RAC_change(mam_inter, 
                             time.var = "year", species.var = "scientificName",
                             abundance.var = "abundance", replicate.var = "siteID")

mam_change_intra <- RAC_change(mam_intra, 
                             time.var = "bout", species.var = "scientificName",
                             abundance.var = "abundance", replicate.var = "siteID") %>%
  #remove bout comparisons that cross years
  filter(substr(bout, 1, 4) == substr(bout2, 1, 4))
```

Fish
```{r}
fish_inter <- readRDS("~/Documents/Projects/temporalNEON/data/fish/fsh_BAT_inter.rds") %>%
  pivot_longer(-c(siteID, year), names_to = "scientificName", values_to = "abundance")

fish_intra <- readRDS("~/Documents/Projects/temporalNEON/data/fish/fsh_BAT_intra.rds") %>%
  pivot_longer(-c(siteID, year, month, bout), names_to = "scientificName", values_to = "abundance")

#sites that match sampling criteria
fish_inc_sites <- fish_intra %>% 
  group_by(siteID) %>% 
  filter(n_distinct(bout) > 2, n_distinct(year) > 3)

fish_change_inter <- RAC_change(fish_inter %>% filter(siteID %in% fish_inc_sites$siteID), 
                             time.var = "year", species.var = "scientificName",
                             abundance.var = "abundance", replicate.var = "siteID")

fish_change_intra <- RAC_change(fish_intra, 
                             time.var = "bout", species.var = "scientificName",
                             abundance.var = "abundance", replicate.var = "siteID") %>%
  #remove bout comparisons that cross years
  filter(substr(bout, 1, 4) == substr(bout2, 1, 4))
```

Inverts
```{r}
invert_inter <- read_csv("~/Documents/Projects/temporalNEON/data/inverts/inverts_table_year_BAT.csv") %>%
  pivot_longer(-c(siteID, year), names_to = "scientificName", values_to = "abundance")

invert_intra <- read_csv("~/Documents/Projects/temporalNEON/data/inverts/inverts_table_bout_BAT.csv") %>%
  group_by(siteID, bout, year, month) %>%
  summarise_all(funs(sum)) %>%
  pivot_longer(-c(siteID, year, month, bout), names_to = "scientificName", values_to = "abundance")

invert_inc_sites <- invert_intra %>% 
  group_by(siteID) %>% 
  filter(n_distinct(bout) > 2, n_distinct(year) > 3)

invert_change_inter <- RAC_change(invert_inter %>% filter(siteID %in% invert_inc_sites$siteID),
                             time.var = "year", species.var = "scientificName",
                             abundance.var = "abundance", replicate.var = "siteID")

invert_change_intra <- RAC_change(invert_intra %>% filter(siteID %in% invert_inc_sites$siteID),
                             time.var = "bout", species.var = "scientificName",
                             abundance.var = "abundance", replicate.var = "siteID")
```

Get all taxa and data in same dataframe
Right now this creates a dataframe that compares intra annual variability to the site average interannual variability
```{r}
library(rgdal)
neon <- readOGR(here::here("Temporal_replication/shapefiles"), layer = "NEON_Field_Sites")
sites <- neon@data

join_taxa_df <- function(df_intra, df_inter){
  change_vars <- c("richness_change", "evenness_change", "rank_change", "gains", "losses")
  taxa_join <- df_intra %>%
    pivot_longer(change_vars, names_to = "variable", values_to = "intra") %>% 
    group_by(siteID, variable) %>%
    mutate(intra_site_mean = mean(na.omit(intra))) %>%
    left_join(df_inter %>% 
                #mutate(year = as.numeric(as.character(year))) %>%
                pivot_longer(change_vars, names_to = "variable", values_to = "inter") %>%
                group_by(siteID, variable) %>% 
                mutate(inter_site_mean = mean(na.omit(inter))) %>%
                select(siteID, variable, inter_site_mean) %>%
                distinct(), 
              by = c("siteID", "variable"))

}

codyn_df <- bind_rows(join_taxa_df(beet_change_intra, beet_change_inter) %>% mutate(taxa = "beetle"),
                      join_taxa_df(mam_change_intra, mam_change_inter) %>% mutate(taxa = "mammal"),
                      join_taxa_df(fish_change_intra, fish_change_inter) %>% mutate(taxa = "fish"),
                      join_taxa_df(invert_change_intra, invert_change_inter) %>% mutate(taxa = "invert")) %>%
  left_join(sites %>% select(siteID = SiteID, latitude = Latitude, longitude = Longitude), by = "siteID")
```

#Plots 

Plot inter vs intra annual variability and intra annual variability by latitude
```{r}
taxa_names <- c("beetle", "mammal", "fish", "invert")

plot_comparisons <- function(taxa_name, x_axis, data = codyn_df){
  change_vars <- c("richness_change", "evenness_change", "rank_change")
  
  ### intra + inter
  # richness_change
  plot <- data %>%
    filter(taxa == taxa_name, variable %in% change_vars) %>%
    ggplot(aes(.data[[x_axis]], intra))+
    geom_point(size=3)+
    geom_smooth(method="glm")+
    geom_point(aes(.data[[x_axis]], intra_site_mean), col="red", size=3)+
    geom_smooth(aes(.data[[x_axis]], intra_site_mean), col="red",method="glm")+
    facet_grid(. ~ variable, scales = "free") +
    xlab(x_axis) +
    theme(panel.background = element_rect(fill = 'white', colour = 'grey85'),
          panel.border = element_rect(fill=NA, colour = "white", size=1),
          axis.line = element_line(color = 'black', size=2.5),
          plot.title = element_text(size=20, vjust=2, family="sans"),
          axis.text.x = element_text(colour='black',size=20),
          axis.text.y = element_text(colour='black',size=20),
          axis.title.x = element_text(colour='black',size=25),
          axis.title.y = element_text(colour='black',size=25),
          axis.ticks = element_line(color = 'black', size=2.5),
          axis.ticks.length=unit(0.5,"cm"),
          legend.position="none",
          strip.text = element_text(size = 15),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank())
  
  mean_plot <- data %>%
    filter(taxa == taxa_name, variable %in% change_vars) %>%
    ggplot(aes(.data[[x_axis]], intra_site_mean))+
    geom_point(col="red", size=3)+
    geom_smooth(col="red",method="glm")+
    facet_grid(. ~ variable, scales = "free") +
    xlab(x_axis) +
    theme(panel.background = element_rect(fill = 'white', colour = 'grey85'),
          panel.border = element_rect(fill=NA, colour = "white", size=1),
          axis.line = element_line(color = 'black', size=2.5),
          plot.title = element_text(size=20, vjust=2, family="sans"),
          axis.text.x = element_text(colour='black',size=20),
          axis.text.y = element_text(colour='black',size=20),
          axis.title.x = element_text(colour='black',size=25),
          axis.title.y = element_text(colour='black',size=25),
          axis.ticks = element_line(color = 'black', size=2.5),
          axis.ticks.length=unit(0.5,"cm"),
          legend.position="none",
          strip.text = element_text(size = 15),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank())
  
  plot_comb <- ggarrange(plot, mean_plot, nrow =2)
  plot_name <- ifelse(x_axis == "inter_site_mean", "-intra-inter.jpeg", "-latitude.jpeg")
  ggsave(file=paste0(here::here("codyn/figures"), "/", taxa_name, plot_name),
         plot_comb, width=15, height=10, dpi=300)
  #width=30,height=10, 
}

map(taxa_names, plot_comparisons, x_axis = "inter_site_mean")
map(taxa_names, plot_comparisons, x_axis = "latitude")
```
