---
title: "NEON download"
author: "Kari Norman"
date: "10/15/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(neonUtilities)
library(piggyback)
```

Download data
```{r}
beetle_dir <- here::here("data", "beetles_raw.rda")
if (file.exists(beetle_dir)){
  load(beetle_dir)
} else{
  beetles_raw <- loadByProduct("DP1.10022.001", check.size = FALSE)
  usethis::use_data(beetles_raw)
}
```

Grab necessary columns from different tables. Important column info:
* sampleID: unique ID for each combination of domainID, siteID, plotID, trapID, and collectDate (unique ID for bet_fielddata table)
* subsampleID: unique ID for each group of individuals, multiple per sample (unique ID for bet_sorting); unclear why there are muliple entries for the same classification
* individualID: unique ID for each individual identified by taxonomists (unique ID for bet_parataxonimist table)

There are multiple levels of taxonomic identification, need to decide which level of rigor we care about (NEON id often just to family, taxonomist ID to species but only for a subset, DNA ID for even smaller subset).

```{r}
data <- beetles_raw$bet_fielddata %>%
  filter(sampleCollected == "Y") %>% #there's an entry for every trap, whether or not they got samples, only want ones with samples
  select(sampleID, domainID, siteID, plotID, trapID, collectDate) %>%
  left_join(beetles_raw$bet_sorting %>%
              filter(sampleType %in% c("carabid", "other carabid")) %>% #only want carabid samples, not bycatch
              select(sampleID, subsampleID, taxonID, scientificName, taxonRank, individualCount,identificationQualifier),
            by = "sampleID") %>%
  filter(!is.na(subsampleID)) %>% #even though they were marked a sampled, some collection times don't acutally have any samples
  left_join(beetles_raw$bet_parataxonomistID %>% select(subsampleID, individualID, taxonID, scientificName, taxonRank,identificationQualifier), by = "subsampleID") %>%
  mutate(year = lubridate::year(collectDate),
         month = lubridate::month(collectDate),
         day = lubridate::day(collectDate))
 
```

```{r}
# Replace sorting taxon info with pinning taxon info (people that pin specimens are more experienced with taxonomy), where available
data_pin <- data %>%
  mutate_if(is.factor, as.character) %>%
  mutate(taxonID = ifelse(is.na(taxonID.y), taxonID.x, taxonID.y)) %>%
  mutate(taxonRank = ifelse(is.na(taxonRank.y), taxonRank.x, taxonRank.y)) %>%
  mutate(scientificName = ifelse(is.na(scientificName.y), scientificName.x, scientificName.y)) %>%
  mutate(identificationSource = ifelse(is.na(scientificName.y), "sort", "pin")) %>%
  mutate (identificationQualifier = ifelse(is.na(taxonID.y), identificationQualifier.x, identificationQualifier.y)) %>%
  select(-ends_with(".x"), -ends_with(".y"))
 
# Add expert taxonomy info, where available
data_pin <- left_join(data_pin, select(beetles_raw$bet_expertTaxonomistIDProcessed,individualID,taxonID,scientificName,taxonRank,identificationQualifier),by = 'individualID')
 
# Replacement old taxon info with expert info, where available
# NOTE - This is repetitive with the code snippet above, and if you want to do it this way you can just combine the calls into one chunk. BUT, you may
#     want to do more than this, as it is *just* a replacement of IDs for individual beetles that an expert identified. If the expert identified
#           a sample as COLSP6 instead of CARSP14, though, then all CARSP14 from that trap on that date should probably be updated to COLSP6…
data_expert <- data_pin %>%
  mutate_if(is.factor, as.character) %>%
  mutate(taxonID = ifelse(is.na(taxonID.y), taxonID.x, taxonID.y)) %>%
  mutate(taxonRank = ifelse(is.na(taxonRank.y), taxonRank.x, taxonRank.y)) %>%
  mutate(scientificName = ifelse(is.na(scientificName.y), scientificName.x, scientificName.y)) %>%
  mutate(identificationSource = ifelse(is.na(scientificName.y), identificationSource, "expert")) %>%
  mutate (identificationQualifier = ifelse(is.na(taxonID.y), identificationQualifier.x, identificationQualifier.y)) %>%
  select(-ends_with(".x"), -ends_with(".y"))
```

What does individual count mean? There are a whole bunch of subsampleID's for which the number of species identified during pinning or by expert taxonomists does not equal the individual count:
```{r}
data_expert %>% 
  group_by(subsampleID, individualCount) %>%
  summarise(n_species = n_distinct(taxonID)) %>% 
  filter(n_species != individualCount) %>%
  dim()
```
For those where only one species was identified, we can probably assume that all the individuals were that species, but when multiple are identified how do we know how individuals are distributed across those species?

How many sites?
How many plots?
How many samples per?
```{r}
data %>%
  select(siteID, plotID, trapID, collectDate, year, month) %>%
  group_by(siteID, plotID, year, month) %>% 
  count() %>% 
  View()
```

What is the distribution of number of samples per plot across years?

What is the distribution of number of samples per trap
```{r}
sample_freq <- data %>%
  select(siteID, plotID, trapID, year, month, collectDate) %>%
  group_by(siteID, plotID, trapID, year) %>% 
  summarise(num_samples = n_distinct(collectDate))

hist(sample_freq$num_samples)
```



